---
title: "Stat 380 Final"
author: "Patrick Erickson, Aiden Shindel, Jordan Brophy"
format: 
  pdf:
    cite-method: biblatex
    pdf-engine: xelatex  
editor: visual
bibliography: references.bib
csl: mla.csl
nocite: |
  @*
appendix: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

# Predicting Drowsy Driving: Can we Machine Learning Model to be able to Predict Drowsiness in Drivers?

## Introduction

### Intuition Behind the Choice of Data Set

-   The data set must be large to be able to get a good testing pool.
-   The data set has meaningful predictors are practically relevant to our test case.
-   The data set contains a drowsy/awake variable
-   The data set follows **F.A.I.R.** and **C.A.R.E.** principles.

### Ethical Considerations in the Choice of our Data Set

In order to ensure the ethical feasibility of the data set, we also scrutinized our selection to follow the **F.A.I.R.** and **C.A.R.E.** principles as discussed in class.

Since we had obtained our data from the open-source data hub Kaggle, we can ensure that anyone can **find** and **access** this data set through Kaggle's use guidelines, stating for a free distribution of data sets that are published for public use to practice ML or data analysis on. The data is also highly versatile, taking a csv format. This is one of the most **interoperable** formats there are for data sets, especially for our use-case in R, where reading in a csv is already built into the base. Lastly, the data set can be **reused** for multiple types of data analysis, and will always remain relevant for its time period.

In terms of **collective benefit**, the data set contributes to an overall understanding of driving without harming any individuals , due to the fact that it is simply raw data. Secondly, as I had stated the author's name, there is clearly ownership over the data set coming from kaggle as mentioned, showing proper **Authority to Control**. Since proper attributions are maintained, it also falls under the **Responsibility and Ethics** guidelines of the C.A.R.E. principles.

The dataset can be found [here](https://www.kaggle.com/datasets/yasharjebraeily/drowsy-detection-dataset?select=Drowsy_datset).

### Cleaning

**Load initial libraries**

```{r}
#|Warning: FALSE
library(png)
library(jpeg)
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(xgboost)

if (!requireNamespace("EBImage", quietly = TRUE)) {
  BiocManager::install("EBImage")
}
library(EBImage)
set.seed(100)
```

Note that the dataset name is Drowsy_datset

```{r}
train_drowsy_dir <- "Drowsy_datset/train/DROWSY"
train_natural_dir <- "Drowsy_datset/train/NATURAL"
test_drowsy_dir <- "Drowsy_datset/test/DROWSY"
test_natural_dir <- "Drowsy_datset/test/NATURAL"

# Check files in each directory
drowsy_files_train <- list.files(train_drowsy_dir, pattern = "\\.png$", full.names = TRUE)
natural_files_train <- list.files(train_natural_dir, pattern = "\\.png$", full.names = TRUE)
drowsy_files_test <- list.files(test_drowsy_dir, pattern = "\\.png$", full.names = TRUE)
natural_files_test <- list.files(test_natural_dir, pattern = "\\.png$", full.names = TRUE)

# Assuming drowsy_files and natural_files are already defined
# For DROWSY files:

# For DROWSY files: Save the first 10 file names in a data frame.
df_drowsy_train <- data.frame(File = head(drowsy_files_train, 5), stringsAsFactors = FALSE)
df_natural_train <- data.frame(File = head(natural_files_train, 5), stringsAsFactors = FALSE)
df_drowsy_test <- data.frame(File = head(drowsy_files_test, 5), stringsAsFactors = FALSE)
df_natural_test <- data.frame(File = head(natural_files_test, 5), stringsAsFactors = FALSE)

# Print the data frames (since they only have 5 rows, printing them prints the head)
kable(df_natural_train, caption = "First 5 NATURAL Train Files:")
kable(df_drowsy_train, caption = "First 5 DROWSY Train Files:")
kable(df_natural_test, caption = "First 5 NATURAL Test Files:")
kable(df_drowsy_test, caption = "First 5 DROWSY Test Files:")

```

This is showing that some files are PNG and some are JPEG.

```{r}
check_png <- function(filepath) {
  tryCatch({
    img <- png::readPNG(filepath)
    return(TRUE)
  }, error = function(e) {
    return(FALSE)
  })
}

# Check a few files with readJPEG:
for (f in drowsy_files_train[1:5]) {
  cat(f, " => PNG check: ", check_png(f), "\n")
}

check_jpeg <- function(filepath) {
  tryCatch({
    jpeg::readJPEG(filepath)
    return(TRUE)
  }, error = function(e) {
    return(FALSE)
  })
}

# Check a few files with readJPEG:
for (f in drowsy_files_train[1:5]) {
  cat(f, " => JPEG check: ", check_jpeg(f), "\n")
}

```

We'll have handle BOTH. The following code chunk loads ALL of the images into a drowsy and natural set, by handling png or jpeg (our data cleaning)

```{r}

# Define a wrapper function to read an image
read_image <- function(filepath) {
  if (check_png(filepath)) {
    img <- png::readPNG(filepath)
  } else if (check_jpeg(filepath)) {
    img <- jpeg::readJPEG(filepath)
  } else {
    stop("Unrecognized image format for file: ", filepath)
  }
  return(img)
}

# We use the function to load all the images as vectors
loaded_images_drowsy_train <- lapply(drowsy_files_train, read_image)
loaded_images_natural_train <- lapply(natural_files_train, read_image)
# We use the function to load all the images as vectors
loaded_images_drowsy_test <- lapply(drowsy_files_test, read_image)
loaded_images_natural_test <- lapply(natural_files_test, read_image)
```

We combine the datasets into one train and test dataset.

```{r}
# Create a base data frame using data.frame()
image_dataset_train <- data.frame(
  image = I(c(loaded_images_drowsy_train, loaded_images_natural_train)),
  label = c(rep("Drowsy", length(loaded_images_drowsy_train)), rep("Natural", length(loaded_images_natural_train))),
  stringsAsFactors = FALSE
)

image_dataset_test <- data.frame(
  image = I(c(loaded_images_drowsy_test, loaded_images_natural_test)),
  label = c(rep("Drowsy", length(loaded_images_drowsy_test)), rep("Natural", length(loaded_images_natural_test))),
  stringsAsFactors = FALSE
)
```

We look at the initial class imbalance. Notice its not that bad. 

```{r}
# For the training dataset:
train_label_counts <- image_dataset_train %>% 
  count(label)
print(train_label_counts)

# For the testing dataset:
test_label_counts <- image_dataset_test %>% 
  count(label)
print(test_label_counts)

# slight imbalance, we will use AUC ROC and f1 scores
```

We shuffle the dataset

```{R}

## Shuffle Datasets
set.seed(100)
image_dataset_train <- image_dataset_train[sample(nrow(image_dataset_train)), ]
image_dataset_test <- image_dataset_test[sample(nrow(image_dataset_test)), ]

```

### Preprocessing

We need to ensure that the datasets are the same going in. This applies CLAHE and HFE to apply pre-processing. 

The first augmentation technique we use therefore is **Contrast Limited Adaptive Histogram Equalization (CLAHE)**. This is a refined version of normal Histogram Equalization that contains a regularization parameter to mitigate the strength of the initial histogram equalization technique by adding a clipping feature to the initial dataset, ensuring that contrasts are never set too high. Furthermore, the image is subdivided into separate sections, where histogram equalization is applied separately. These sections are then combined using bilinear interpolation, which ensures a smooth transition between each segmented histogram. This way, we do not create entire areas of extreme contrast and intensity, something that normal histogram equalization is prone to doing. We therefore opted to maintain this augmentation technique in our pipeline.


Lastly, **High-Frequency Emphasis (HFE)** is a type of processing technique that boosts subtle details and fine textured by enhancing the frequency components of a given image. The "frequency" is usually represented via a Fourier Transform into many different frequency elements, where low frequencies correlated to slowly varying information (think features that are fairly consistent throughout an image) and high frequencies relating to contrasting sections of an image (areas where features abruptly change). This allows for the convolutions in computer vision to better pick up patterns and improve generalizability. After carefully analyzing our own parameters and the differences between our and our parents' methodologies, we opted to maintain this data processing technique, as it is vital for the extrapolation of the plant diseases. For our high frequency emphasis, we use the following Gaussian blur for convolutions with a sliding of 1. (THIS IS NOT DEEP LEARNING, SIMPLY A TRANSFORMATION WITH STATIC CONVOLVING FILTERS, NOT LEARNABLE FILTERS).
$$
\frac{1}{10}
\begin{pmatrix}
1 & 2 & 1 \\
2 & 4 & 2 \\
1 & 2 & 1
\end{pmatrix}
$$

```{R}
applyFeatureEngineering <- function(df) {
  
  # Helper function to apply CLAHE using EBImage.
  apply_CLAHE <- function(img_matrix) {
    # Convert the 48x48 matrix into an EBImage object.
    img <- EBImage::Image(img_matrix)
    
    # Apply CLAHE.
    # Note: Your version of EBImage must have the clahe() function available.
    img_clahe <- EBImage::clahe(img, nx = 8, ny = 8, bins = 256, limit = 2, keep.range = FALSE)

    
    # Convert back to a matrix.
    return(as.matrix(img_clahe))
  }
  
  # Helper function for High Frequency Extraction (HFE).
  high_frequency_extraction <- function(img_matrix) {
    # Convert the 48x48 matrix to an EBImage object.
    img <- EBImage::Image(img_matrix)
    
    # Apply a Gaussian blur to capture low frequency details (sigma can be adjusted).
    img_blurred <- EBImage::gblur(img, sigma = 1)
    
    # Subtract the blurred image from the original image to extract high frequencies.
    img_hfe <- img - img_blurred
    
    # Convert the result back to a matrix.
    return(as.matrix(img_hfe))
  }
  
  # Apply the feature engineering functions to each image in the dataset.
  # This creates two new columns:
  #   - image_clahe: with CLAHE-enhanced images.
  #   - image_hfe: with high frequency extracted images.
  df_new <- df %>%
    mutate(
      image = map(image, ~ apply_CLAHE(.))
    ) %>%
    mutate(
      image = map(image, ~ high_frequency_extraction(.))
    )
  
  return(df_new)
}

```

```{r}
processedtrain <- applyFeatureEngineering(image_dataset_train)
processedTest <- applyFeatureEngineering(image_dataset_test)

```

### Readying for models

We define our new models as one with feature engineering and one without feature engineering. We stretch and flatten our images into one long string of vectors.

```{r}
trainNoFeatureEngineering <- image_dataset_train %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")

testNoFeatureEngineering <- image_dataset_test %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")
```

```{r}
trainWithFeatureEngineering <- processedtrain %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")

testWithFeatureEngineering <- processedTest %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")
```

These are plotting functions. This one reconstructs the image from the flattened array.

```{r}
# We'll assume your expanded data frame is called train_df_expanded.
# Each row has a "label" column plus columns flat_image_1, flat_image_2, ..., flat_image_2304.

reconstruct_image <- function(df_row, label_col = "label") {

  label_value <- df_row[[label_col]]
  pixel_columns <- grep("^flat_image_", names(df_row), value = TRUE)
  pixel_values  <- as.numeric(df_row[pixel_columns])
  mat <- matrix(pixel_values, nrow = 48, ncol = 48, byrow = TRUE)
  
  image(
    x      = 1:48,
    y      = 1:48,
    z      = t(apply(mat, 2, rev)),
    col    = gray(seq(0, 1, length.out = 256)),
    main   = paste("Label:", label_value),
    xlab   = "",
    ylab   = "",
    axes   = FALSE,
    asp    = 1
  )
  
  return(mat)
}

```

This one plots the image.

```{r}

# Function that plots the feature flattened images for 
plot_image_ggplot <- function(df_row, label_col = "label") {
  label_value <- df_row[[label_col]]
  
  pixel_columns <- grep("^flat_image_", names(df_row), value = TRUE)
  pixel_values  <- as.numeric(df_row[pixel_columns])
  
  image_matrix <- matrix(pixel_values, nrow = 48, ncol = 48, byrow = TRUE)
  
  df_image <- expand.grid(x = 1:48, y = 1:48)
  
  df_image$fill <- as.vector(t(image_matrix))
  
  p <- ggplot(df_image, aes(x = x, y = y, fill = fill)) +
    geom_raster() +
    scale_fill_gradient(low = "black", high = "white") +
    scale_y_reverse() +
    labs(title = paste("Label:", label_value)) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      plot.title = element_text(hjust = 0.5)
    )
  
  return(p)
}


```

We print the first 9. Notice the differences.

```{r}

# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(trainNoFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)


```

```{r}

# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(testNoFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)

```

```{r}
# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(trainWithFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)
```

We need to change the scaling back to \[0,1\] after feature engineering.

```{r}
trainWithFeatureEngineering <- trainWithFeatureEngineering %>%
  mutate(across(where(is.numeric), ~ (. - min(.)) / (max(.) - min(.))))
testWithFeatureEngineering <- testWithFeatureEngineering %>%
  mutate(across(where(is.numeric), ~ (. - min(.)) / (max(.) - min(.))))
```

```{r}
# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(trainWithFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)
```

```{r}
# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(testWithFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)
```


Save our datasets here.
```{r}
write.csv(trainNoFeatureEngineering, file = "Prepped_Data/trainNoFeatureEngineering.csv", row.names = FALSE)
write.csv(testNoFeatureEngineering, file = "Prepped_Data/testNoFeatureEngineering.csv", row.names = FALSE)
write.csv(trainWithFeatureEngineering, file = "Prepped_Data/trainWithFeatureEngineering.csv", row.names = FALSE)
write.csv(testWithFeatureEngineering, file = "Prepped_Data/testWithFeatureEngineering.csv", row.names = FALSE)
```

### Do PCA

we will do PCA. We will use the PCA done on the train set to transform our test set to then evaluate with our models later.

### For no pre-processing

```{r}

# It’s common to remove non-numeric columns when performing PCA.
# Here we automatically select only numeric columns.
numeric_data <- trainNoFeatureEngineering %>% 
  select(where(is.numeric))

# Perform PCA, scaling the variables since they may be on different scales
pca_result <- prcomp(numeric_data, scale. = TRUE)

# Calculate the proportion of variance explained for each principal component.
# pca_result$sdev holds the standard deviations for each PC.
variance_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cumulative_variance <- cumsum(variance_explained)

# Create a data frame for plotting
pca_df <- data.frame(
  PC = 1:length(variance_explained),
  VarianceExplained = variance_explained,
  CumulativeVariance = cumulative_variance
)

# Optionally, determine the number of components needed to explain at least 95% of the variance.
target_threshold <- 0.95
n_components <- which(cumulative_variance >= target_threshold)[1]
cat(sprintf("Number of components to reach at least %.0f%% variance explained: %d", target_threshold * 100, n_components))

```

```{r}
# Create a scree plot with both the bar chart (variance explained per PC) and a line for cumulative variance.
scree_plot <- ggplot(pca_df, aes(x = PC)) +
  geom_bar(aes(y = VarianceExplained), stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = CumulativeVariance), color = "red", size = 1) +
  geom_point(aes(y = CumulativeVariance), color = "red", size = 2) +
  geom_vline(xintercept = n_components, linetype = "dashed", color = "darkgreen") +
  labs(title = "Scree Plot for PCA",
       subtitle = sprintf("Vertical dashed line indicates %d components (>= %.0f%% variance explained)", n_components, target_threshold*100),
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  theme_minimal()

# Display the plot
print(scree_plot)

```

```{r}

test_numeric <- testNoFeatureEngineering %>%
  select(-label) %>%
  select(where(is.numeric))

# For the training set
pca_train <- as.data.frame(pca_result$x[, 1:183])
pca_train$label <- trainNoFeatureEngineering$label
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))

# For the test set:
scaled_test <- scale(test_numeric, center = pca_result$center, scale = pca_result$scale)
pca_test_full <- as.data.frame(as.matrix(scaled_test) %*% pca_result$rotation)
pca_test <- pca_test_full[, 1:183]  
pca_test$label <- testNoFeatureEngineering$label
pca_test$label <- factor(pca_test$label, levels = c("Drowsy", "Natural"))
```

```{r}
write.csv(pca_train, file = "Prepped_Data/trainNoFeatureEngineeringPCA.csv", row.names = FALSE)
write.csv(pca_test, file = "Prepped_Data/testNoFeatureEngineeringPCA.csv", row.names = FALSE)
```


For pre-processed, we also perform pca.

```{r}
# Here we automatically select only numeric columns.
numeric_data <- trainWithFeatureEngineering %>% 
  select(where(is.numeric))

# Perform PCA, scaling the variables since they may be on different scales
pca_result <- prcomp(numeric_data, scale. = TRUE)

# Calculate the proportion of variance explained for each principal component.
# pca_result$sdev holds the standard deviations for each PC.
variance_explained <- pca_result$sdev^2 / sum(pca_result$sdev^2)
cumulative_variance <- cumsum(variance_explained)

# Create a data frame for plotting
pca_df <- data.frame(
  PC = 1:length(variance_explained),
  VarianceExplained = variance_explained,
  CumulativeVariance = cumulative_variance
)

```

```{r}
# Optionally, determine the number of components needed to explain at least 90% of the variance.
target_threshold <- 0.95
n_components <- which(cumulative_variance >= target_threshold)[1]
cat(sprintf("Number of components to reach at least %.0f%% variance explained: %d", target_threshold * 100, n_components))

# Create a scree plot with both the bar chart (variance explained per PC) and a line for cumulative variance.
scree_plot <- ggplot(pca_df, aes(x = PC)) +
  geom_bar(aes(y = VarianceExplained), stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(y = CumulativeVariance), color = "red", size = 1) +
  geom_point(aes(y = CumulativeVariance), color = "red", size = 2) +
  geom_vline(xintercept = n_components, linetype = "dashed", color = "darkgreen") +
  labs(title = "Scree Plot for PCA",
       subtitle = sprintf("Vertical dashed line indicates %d components (>= %.0f%% variance explained)", n_components, target_threshold*100),
       x = "Principal Component",
       y = "Proportion of Variance Explained") +
  theme_minimal()

# Display the plot
print(scree_plot)
```


The contrast actually increases the variance a great deal. While this might be helpful for complex models, for simple ML models it could hurt it. 
```{r}
test_numeric <- testWithFeatureEngineering %>%
  select(-label) %>%
  select(where(is.numeric))

# For the training set
pca_train <- as.data.frame(pca_result$x[, 1:1313])
pca_train$label <- trainWithFeatureEngineering$label
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))

# For the test set:
scaled_test <- scale(test_numeric, center = pca_result$center, scale = pca_result$scale)
pca_test_full <- as.data.frame(as.matrix(scaled_test) %*% pca_result$rotation)
pca_test <- pca_test_full[, 1:1313]  # subset to the first 183 principal components
pca_test$label <- testWithFeatureEngineering$label
pca_test$label <- factor(pca_test$label, levels = c("Drowsy", "Natural"))

```

```{r}
write.csv(pca_train, file = "Prepped_Data/trainWithFeatureEngineeringPCA.csv", row.names = FALSE)
write.csv(pca_test, file = "Prepped_Data/testWithFeatureEngineeringPCA.csv", row.names = FALSE)
```

### Model Ablation Study
I didn't take out the libraries. Its more important to run the models right now.All of these have their own seperate hyperparameter search and are 5 fold cv'd. We save a confusion matrix, some metrics like accuracy, recall, precision, f1 score, AUC from ROC, the actual ROC data points, and a png of the ggplot of the AUC ROC per model. The actual ROC data points can be used to display multiple model AUCs on one plot, if we so choose to later. Lastly, we save the models as .rdl. We will need this to open them back up for our stacking model.

Gaussian SVM, No feature pre-processing, PCA

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainNoFeatureEngineeringPCA.csv", show_col_types = FALSE)
pca_test <- read_csv("Prepped_Data/testNoFeatureEngineeringPCA.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# If necessary, reorder levels so that "Drowsy" is the positive class.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# Note: We add classProbs = TRUE and summaryFunction for ROC
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an SVM model.
# The additional argument prob.model = TRUE ensures that ksvm computes probabilities.
svm_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "svmRadial", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy",   # Optimize model using Accuracy
  prob.model = TRUE # Ensure probability estimates are computed
)

# Stop the parallel cluster after training
stopCluster(cl)

# Display model results
print(svm_model$results)
print(svm_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(svm_model, file = "Base_Models/SVMModelNoFeatureEngineeringPCA.rds")
cat("Model saved as 'Base_Models/SVMModelNoFeatureEngineeringPCA.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/SVMModelNoFeatureEngineeringPCA.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "SVMNoFeatureEngineeringPCA")
###########################################
best_tune_df <- cbind(model = "SVMNoFeatureEngineeringPCA", svm_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/SVM/BestHyperparams_SVMNoFeatureEngineeringPCA.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/SVM/BestHyperparams_SVMNoFeatureEngineeringPCA.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(svm_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "SVMNoFeatureEngineeringPCA",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/SVM/ConfusionMatrix_SVMNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/SVM/ConfusionMatrix_SVMNoFeatureEngineeringPCA.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (now that we enabled probabilities)
  probs <- predict(svm_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "SVMNoFeatureEngineeringPCA",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/SVM/Metrics_SVMNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/SVM/Metrics_SVMNoFeatureEngineeringPCA.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/SVM/ROCData_SVMNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/SVM/ROCData_SVMNoFeatureEngineeringPCA.csv'.\n")
  
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for SVMNoFeatureEngineeringPCA") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/SVM/ROC_SVMNoFeatureEngineeringPCA.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/SVM/ROC_SVMNoFeatureEngineeringPCA.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

XGBoost, No feature pre-processing, PCA

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
library(xgboost)  # For the xgbTree method
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainNoFeatureEngineeringPCA.csv", show_col_types = FALSE)
pca_test <- read_csv("Prepped_Data/testNoFeatureEngineeringPCA.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" is the positive class.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to get AUC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an XGBoost model using caret's xgbTree method
xgb_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "xgbTree", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using Accuracy
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(xgb_model$results)
print(xgb_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(xgb_model, file = "Base_Models/XGBoostNoFeatureEngineeringPCA.rds")
cat("Model saved as 'Base_Models/XGBoostNoFeatureEngineeringPCA.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/XGBoostNoFeatureEngineeringPCA.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "XGBoostNoFeatureEngineeringPCA")
###########################################
best_tune_df <- cbind(model = "XGBoostNoFeatureEngineeringPCA", xgb_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/XGBoost/BestHyperparams_XGBoostNoFeatureEngineeringPCA.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/XGBoost/BestHyperparams_XGBoostNoFeatureEngineeringPCA.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(xgb_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "XGBoostNoFeatureEngineeringPCA",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostNoFeatureEngineeringPCA.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (ensure type = "prob")
  probs <- predict(xgb_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "XGBoostNoFeatureEngineeringPCA",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/XGBoost/Metrics_XGBoostNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/XGBoost/Metrics_XGBoostNoFeatureEngineeringPCA.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/XGBoost/ROCData_XGBoostNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/XGBoost/ROCData_XGBoostNoFeatureEngineeringPCA.csv'.\n")
  
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for XGBoostNoFeatureEngineeringPCA") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/XGBoost/ROC_XGBoostNoFeatureEngineeringPCA.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/XGBoost/ROC_XGBoostNoFeatureEngineeringPCA.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

RandomForest, No feature pre-processing, PCA

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Random Forest can be run using caret, so an explicit load of randomForest is optional.
# library(randomForest)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainNoFeatureEngineeringPCA.csv", show_col_types = FALSE)
pca_test  <- read_csv("Prepped_Data/testNoFeatureEngineeringPCA.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" (the positive class) is first.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to obtain ROC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train a Random Forest model using caret's "rf" method
rf_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "rf", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using ROC
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(rf_model$results)
print(rf_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(rf_model, file = "Base_Models/RandomForestNoFeatureEngineeringPCA.rds")
cat("Model saved as 'Base_Models/RandomForestNoFeatureEngineeringPCA.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/RandomForestNoFeatureEngineeringPCA.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "RandomForestNoFeatureEngineeringPCA")
###########################################
best_tune_df <- cbind(model = "RandomForestNoFeatureEngineeringPCA", rf_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/Random_Forest/BestHyperparams_RandomForestNoFeatureEngineeringPCA.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/Random_Forest/BestHyperparams_RandomForestNoFeatureEngineeringPCA.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(rf_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix values; assumes binary classification.
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "RandomForestNoFeatureEngineeringPCA",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestNoFeatureEngineeringPCA.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy  <- conf_matrix$overall["Accuracy"]
  recall    <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score  <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (type = "prob")
  probs <- predict(rf_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "RandomForestNoFeatureEngineeringPCA",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/Random_Forest/Metrics_RandomForestNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/Random_Forest/Metrics_RandomForestNoFeatureEngineeringPCA.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build ROC data: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/Random_Forest/ROCData_RandomForestNoFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("ROC data saved as 'Base_Models_Data/Random_Forest/ROCData_RandomForestNoFeatureEngineeringPCA.csv'.\n")
  
  # Generate the ROC plot using ggplot2
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # 45° dotted reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for RandomForestNoFeatureEngineeringPCA") +
    theme_minimal() +
    # Annotate AUC in the top right corner (rounded to 3 decimals)
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file
  ggsave("Base_Models_Data/Random_Forest/ROC_RandomForestNoFeatureEngineeringPCA.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/Random_Forest/ROC_RandomForestNoFeatureEngineeringPCA.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

Gaussian SVM, feature pre-processing, PCA

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainWithFeatureEngineeringPCA.csv", show_col_types = FALSE)
pca_test <- read_csv("Prepped_Data/testWithFeatureEngineeringPCA.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# If necessary, reorder levels so that "Drowsy" is the positive class.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# Note: We add classProbs = TRUE and summaryFunction for ROC
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an SVM model.
# The additional argument prob.model = TRUE ensures that ksvm computes probabilities.
svm_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "svmRadial", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy",   # Optimize model using ROC
  prob.model = TRUE # Ensure probability estimates are computed
)

# Stop the parallel cluster after training
stopCluster(cl)

# Display model results
print(svm_model$results)
print(svm_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(svm_model, file = "Base_Models/SVMModelWithFeatureEngineeringPCA.rds")
cat("Model saved as 'Base_Models/SVMModelWithFeatureEngineeringPCA.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/SVMModelWithFeatureEngineeringPCA.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "SVMWithFeatureEngineeringPCA")
###########################################
best_tune_df <- cbind(model = "SVMWithFeatureEngineeringPCA", svm_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/SVM/BestHyperparams_SVMWithFeatureEngineeringPCA.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/SVM/BestHyperparams_SVMWithFeatureEngineeringPCA.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(svm_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "SVMWithFeatureEngineeringPCA",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/SVM/ConfusionMatrix_SVMWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/SVM/ConfusionMatrix_SVMWithFeatureEngineeringPCA.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (now that we enabled probabilities)
  probs <- predict(svm_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "SVMWithFeatureEngineeringPCA",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/SVM/Metrics_SVMWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/SVM/Metrics_SVMWithFeatureEngineeringPCA.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/SVM/ROCData_SVMWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/SVM/ROCData_SVMWithFeatureEngineeringPCA.csv'.\n")
  
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for SVMWithFeatureEngineeringPCA") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/SVM/ROC_SVMWithFeatureEngineeringPCA.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/SVM/ROC_SVMWithFeatureEngineeringPCA.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

XGBoost, feature pre-processing, PCA

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
library(xgboost)  # For the xgbTree method
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainWithFeatureEngineeringPCA.csv", show_col_types = FALSE)
pca_test <- read_csv("Prepped_Data/testWithFeatureEngineeringPCA.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" is the positive class.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to get AUC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an XGBoost model using caret's xgbTree method
xgb_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "xgbTree", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using ROC
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(xgb_model$results)
print(xgb_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(xgb_model, file = "Base_Models/XGBoostWithFeatureEngineeringPCA.rds")
cat("Model saved as 'Base_Models/XGBoostWithFeatureEngineeringPCA.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/XGBoostWithFeatureEngineeringPCA.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "XGBoostWithFeatureEngineeringPCA")
###########################################
best_tune_df <- cbind(model = "XGBoostWithFeatureEngineeringPCA", xgb_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/XGBoost/BestHyperparams_XGBoostWithFeatureEngineeringPCA.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/XGBoost/BestHyperparams_XGBoostWithFeatureEngineeringPCA.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(xgb_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "XGBoostWithFeatureEngineeringPCA",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostWithFeatureEngineeringPCA.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (ensure type = "prob")
  probs <- predict(xgb_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "XGBoostWithFeatureEngineeringPCA",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/XGBoost/Metrics_XGBoostWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/XGBoost/Metrics_XGBoostWithFeatureEngineeringPCA.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/XGBoost/ROCData_XGBoostWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/XGBoost/ROCData_XGBoostWithFeatureEngineeringPCA.csv'.\n")
  
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for XGBoostWithFeatureEngineeringPCA") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/XGBoost/ROC_XGBoostWithFeatureEngineeringPCA.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/XGBoost/ROC_XGBoostWithFeatureEngineeringPCA.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

RandomForest, feature pre-processing, PCA


```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Random Forest can be run using caret, so an explicit load of randomForest is optional.
# library(randomForest)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainWithFeatureEngineeringPCA.csv", show_col_types = FALSE)
pca_test  <- read_csv("Prepped_Data/testWithFeatureEngineeringPCA.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" (the positive class) is first.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to obtain ROC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train a Random Forest model using caret's "rf" method
rf_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "rf", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using ROC
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(rf_model$results)
print(rf_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(rf_model, file = "Base_Models/RandomForestWithFeatureEngineeringPCA.rds")
cat("Model saved as 'Base_Models/RandomForestWithFeatureEngineeringPCA.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/RandomForestWithFeatureEngineeringPCA.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "RandomForestWithFeatureEngineeringPCA")
###########################################
best_tune_df <- cbind(model = "RandomForestWithFeatureEngineeringPCA", rf_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/Random_Forest/BestHyperparams_RandomForestWithFeatureEngineeringPCA.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/Random_Forest/BestHyperparams_RandomForestWithFeatureEngineeringPCA.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(rf_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix values; assumes binary classification.
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "RandomForestWithFeatureEngineeringPCA",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestWithFeatureEngineeringPCA.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy  <- conf_matrix$overall["Accuracy"]
  recall    <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score  <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (type = "prob")
  probs <- predict(rf_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "RandomForestWithFeatureEngineeringPCA",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/Random_Forest/Metrics_RandomForestWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/Random_Forest/Metrics_RandomForestWithFeatureEngineeringPCA.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build ROC data: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/Random_Forest/ROCData_RandomForestWithFeatureEngineeringPCA.csv", row.names = FALSE)
  cat("ROC data saved as 'Base_Models_Data/Random_Forest/ROCData_RandomForestWithFeatureEngineeringPCA.csv'.\n")
  
  # Generate the ROC plot using ggplot2
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # 45° dotted reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for RandomForestWithFeatureEngineeringPCA") +
    theme_minimal() +
    # Annotate AUC in the top right corner (rounded to 3 decimals)
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file
  ggsave("Base_Models_Data/Random_Forest/ROC_RandomForestWithFeatureEngineeringPCA.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/Random_Forest/ROC_RandomForestWithFeatureEngineeringPCA.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

Gaussian SVM, No feature pre-processing


```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
train <- read_csv("Prepped_Data/trainNoFeatureEngineering.csv", show_col_types = FALSE)
test <- read_csv("Prepped_Data/testNoFeatureEngineering.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# If necessary, reorder levels so that "Drowsy" is the positive class.
train$label <- factor(train$label, levels = c("Drowsy", "Natural"))
test$label  <- factor(test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# Note: We add classProbs = TRUE and summaryFunction for ROC
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an SVM model.
# The additional argument prob.model = TRUE ensures that ksvm computes probabilities.
svm_model <- train(
  label ~ ., 
  data = train, 
  method = "svmRadial", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy",   # Optimize model using ROC
  prob.model = TRUE # Ensure probability estimates are computed
)

# Stop the parallel cluster after training
stopCluster(cl)

# Display model results
print(svm_model$results)
print(svm_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(svm_model, file = "Base_Models/SVMModelNoFeatureEngineering.rds")
cat("Model saved as 'Base_Models/SVMModelNoFeatureEngineering.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/SVMModelNoFeatureEngineering.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "SVMNoFeatureEngineering")
###########################################
best_tune_df <- cbind(model = "SVMNoFeatureEngineering", svm_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/SVM/BestHyperparams_SVMNoFeatureEngineering.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/SVM/BestHyperparams_SVMNoFeatureEngineering.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(svm_model, newdata = test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "SVMNoFeatureEngineering",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/SVM/ConfusionMatrix_SVMNoFeatureEngineering.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/SVM/ConfusionMatrix_SVMNoFeatureEngineering.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (now that we enabled probabilities)
  probs <- predict(svm_model, newdata = test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "SVMNoFeatureEngineering",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/SVM/Metrics_SVMNoFeatureEngineering.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/SVM/Metrics_SVMNoFeatureEngineering.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  write.csv(roc_data, "Base_Models_Data/SVM/ROCData_SVMNoFeatureEngineering.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/SVM/ROCData_SVMNoFeatureEngineering.csv'.\n")
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for SVMNoFeatureEngineering") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/SVM/ROC_SVMNoFeatureEngineering.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/SVM/ROC_SVMNoFeatureEngineering.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

XGBoost, No feature pre-processing


```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
library(xgboost)  # For the xgbTree method
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainNoFeatureEngineering.csv", show_col_types = FALSE)
pca_test <- read_csv("Prepped_Data/testNoFeatureEngineering.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" is the positive class.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to get AUC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an XGBoost model using caret's xgbTree method
xgb_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "xgbTree", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using ROC
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(xgb_model$results)
print(xgb_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(xgb_model, file = "Base_Models/XGBoostNoFeatureEngineering.rds")
cat("Model saved as 'Base_Models/XGBoostNoFeatureEngineering.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/XGBoostNoFeatureEngineering.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "XGBoostNoFeatureEngineering")
###########################################
best_tune_df <- cbind(model = "XGBoostNoFeatureEngineering", xgb_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/XGBoost/BestHyperparams_XGBoostNoFeatureEngineering.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/XGBoost/BestHyperparams_XGBoostNoFeatureEngineering.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(xgb_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "XGBoostNoFeatureEngineering",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostNoFeatureEngineering.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostNoFeatureEngineering.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (ensure type = "prob")
  probs <- predict(xgb_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "XGBoostNoFeatureEngineering",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/XGBoost/Metrics_XGBoostNoFeatureEngineering.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/XGBoost/Metrics_XGBoostNoFeatureEngineering.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/XGBoost/ROCData_XGBoostNoFeatureEngineering.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/XGBoost/ROCData_XGBoostNoFeatureEngineering.csv'.\n")
  
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for XGBoostNoFeatureEngineering") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/XGBoost/ROC_XGBoostNoFeatureEngineering.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/XGBoost/ROC_XGBoostNoFeatureEngineering.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

Random Forest, No feature pre-processing


```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Random Forest can be run using caret, so an explicit load of randomForest is optional.
# library(randomForest)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainNoFeatureEngineering.csv", show_col_types = FALSE)
pca_test  <- read_csv("Prepped_Data/testNoFeatureEngineering.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" (the positive class) is first.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to obtain ROC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train a Random Forest model using caret's "rf" method
rf_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "rf", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using ROC
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(rf_model$results)
print(rf_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(rf_model, file = "Base_Models/RandomForestNoFeatureEngineering.rds")
cat("Model saved as 'Base_Models/RandomForestNoFeatureEngineering.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/RandomForestNoFeatureEngineering.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "RandomForestNoFeatureEngineering")
###########################################
best_tune_df <- cbind(model = "RandomForestNoFeatureEngineering", rf_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/Random_Forest/BestHyperparams_RandomForestNoFeatureEngineering.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/Random_Forest/BestHyperparams_RandomForestNoFeatureEngineering.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(rf_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix values; assumes binary classification.
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "RandomForestNoFeatureEngineering",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestNoFeatureEngineering.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestNoFeatureEngineering.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy  <- conf_matrix$overall["Accuracy"]
  recall    <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score  <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (type = "prob")
  probs <- predict(rf_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "RandomForestNoFeatureEngineering",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/Random_Forest/Metrics_RandomForestNoFeatureEngineering.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/Random_Forest/Metrics_RandomForestNoFeatureEngineering.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build ROC data: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/Random_Forest/ROCData_RandomForestNoFeatureEngineering.csv", row.names = FALSE)
  cat("ROC data saved as 'Base_Models_Data/Random_Forest/ROCData_RandomForestNoFeatureEngineering.csv'.\n")
  
  # Generate the ROC plot using ggplot2
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # 45° dotted reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for RandomForestNoFeatureEngineering") +
    theme_minimal() +
    # Annotate AUC in the top right corner (rounded to 3 decimals)
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file
  ggsave("Base_Models_Data/Random_Forest/ROC_RandomForestNoFeatureEngineering.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/Random_Forest/ROC_RandomForestNoFeatureEngineering.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

Gaussian SVM, feature pre-processing

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
train <- read_csv("Prepped_Data/trainWithFeatureEngineering.csv", show_col_types = FALSE)
test <- read_csv("Prepped_Data/testWithFeatureEngineering.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# If necessary, reorder levels so that "Drowsy" is the positive class.
train$label <- factor(train$label, levels = c("Drowsy", "Natural"))
test$label  <- factor(test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# Note: We add classProbs = TRUE and summaryFunction for ROC
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an SVM model.
# The additional argument prob.model = TRUE ensures that ksvm computes probabilities.
svm_model <- train(
  label ~ ., 
  data = train, 
  method = "svmRadial", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy",   # Optimize model using ROC
  prob.model = TRUE # Ensure probability estimates are computed
)

# Stop the parallel cluster after training
stopCluster(cl)

# Display model results
print(svm_model$results)
print(svm_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(svm_model, file = "Base_Models/SVMModelWithFeatureEngineering.rds")
cat("Model saved as 'Base_Models/SVMModelWithFeatureEngineering.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/SVMModelWithFeatureEngineering.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "SVMWithFeatureEngineering")
###########################################
best_tune_df <- cbind(model = "SVMWithFeatureEngineering", svm_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/SVM/BestHyperparams_SVMWithFeatureEngineering.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/SVM/BestHyperparams_SVMWithFeatureEngineering.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(svm_model, newdata = test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "SVMWithFeatureEngineering",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/SVM/ConfusionMatrix_SVMWithFeatureEngineering.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/SVM/ConfusionMatrix_SVMWithFeatureEngineering.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (now that we enabled probabilities)
  probs <- predict(svm_model, newdata = test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "SVMWithFeatureEngineering",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/SVM/Metrics_SVMWithFeatureEngineering.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/SVM/Metrics_SVMWithFeatureEngineering.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/SVM/ROCData_SVMWithFeatureEngineering.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/SVM/ROCData_SVMWithFeatureEngineering.csv'.\n")
  
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for SVMWithFeatureEngineering") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/SVM/ROC_SVMWithFeatureEngineering.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/SVM/ROC_SVMWithFeatureEngineering.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

XGBoost, feature pre-processing

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
library(xgboost)  # For the xgbTree method
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainWithFeatureEngineeringPCA.csv", show_col_types = FALSE)
pca_test <- read_csv("Prepped_Data/testWithFeatureEngineeringPCA.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" is the positive class.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to get AUC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train an XGBoost model using caret's xgbTree method
xgb_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "xgbTree", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using ROC
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(xgb_model$results)
print(xgb_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(xgb_model, file = "Base_Models/XGBoostWithFeatureEngineering.rds")
cat("Model saved as 'Base_Models/XGBoostWithFeatureEngineering.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/XGBoostWithFeatureEngineering.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "XGBoostWithFeatureEngineering")
###########################################
best_tune_df <- cbind(model = "XGBoostWithFeatureEngineering", xgb_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/XGBoost/BestHyperparams_XGBoostWithFeatureEngineering.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/XGBoost/BestHyperparams_XGBoostWithFeatureEngineering.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(xgb_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix table; assumes binary classification.
# Here we define the first level as the positive class ("Drowsy").
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "XGBoostWithFeatureEngineering",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostWithFeatureEngineering.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/XGBoost/ConfusionMatrix_XGBoostWithFeatureEngineering.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy <- conf_matrix$overall["Accuracy"]
  recall <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (ensure type = "prob")
  probs <- predict(xgb_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "XGBoostWithFeatureEngineering",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/XGBoost/Metrics_XGBoostWithFeatureEngineering.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/XGBoost/Metrics_XGBoostWithFeatureEngineering.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build data for ROC curve: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/XGBoost/ROCData_XGBoostWithFeatureEngineering.csv", row.names = FALSE)
  cat("ROC Data for plot building saved as 'Base_Models_Data/XGBoost/ROCData_XGBoostWithFeatureEngineering.csv'.\n")
  
  # Generate the ROC plot using ggplot2.
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # Dotted 45° reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for XGBoostWithFeatureEngineering") +
    theme_minimal() +
    # Annotate AUC in the top right corner, rounded to 3 significant figures.
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file.
  ggsave("Base_Models_Data/XGBoost/ROC_XGBoostWithFeatureEngineering.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/XGBoost/ROC_XGBoostWithFeatureEngineering.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```

Random Forest, feature pre-processing


```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(kernlab)
library(BiocManager)
library(readr)
library(doParallel)
library(EBImage)
# Random Forest can be run using caret, so an explicit load of randomForest is optional.
# library(randomForest)
# Added for ROC analysis
library(pROC)

set.seed(100)

# Load training and testing data
pca_train <- read_csv("Prepped_Data/trainWithFeatureEngineering.csv", show_col_types = FALSE)
pca_test  <- read_csv("Prepped_Data/testWithFeatureEngineering.csv", show_col_types = FALSE)

# Ensure that the target variable is a factor.
# Reorder levels so that "Drowsy" (the positive class) is first.
pca_train$label <- factor(pca_train$label, levels = c("Drowsy", "Natural"))
pca_test$label  <- factor(pca_test$label,  levels = c("Drowsy", "Natural"))

# Set up parallel processing
num_cores <- parallel::detectCores() - 1  
cl <- makeCluster(num_cores)
registerDoParallel(cl)

# Define cross-validation method for training.
# We add classProbs = TRUE and summaryFunction = defaultSummary to obtain ROC and probability estimates.
train_control <- trainControl(
  method = "cv",
  number = 5,
  allowParallel = TRUE,
  classProbs = TRUE,
  summaryFunction = defaultSummary
)

# Train a Random Forest model using caret's "rf" method
rf_model <- train(
  label ~ ., 
  data = pca_train, 
  method = "rf", 
  trControl = train_control,
  preProcess = NULL,
  metric = "Accuracy"  # Optimize model using ROC
)

# Stop the parallel cluster after training
stopCluster(cl)

# Print the results and best tuning parameters
print(rf_model$results)
print(rf_model$bestTune)

###########################################
# 1. Save the model and show how to call it later
###########################################
saveRDS(rf_model, file = "Base_Models/RandomForestWithFeatureEngineering.rds")
cat("Model saved as 'Base_Models/RandomForestWithFeatureEngineering.rds'. To load it later, use:\n")
cat("loaded_model <- readRDS('Base_Models/RandomForestWithFeatureEngineering.rds')\n\n")

###########################################
# 2. Save bestTune hyperparameters as CSV
#    (include a column 'model' with the label "RandomForestWithFeatureEngineering")
###########################################
best_tune_df <- cbind(model = "RandomForestWithFeatureEngineering", rf_model$bestTune)
write.csv(best_tune_df, "Base_Models_Data/Random_Forest/BestHyperparams_RandomForestWithFeatureEngineering.csv", row.names = FALSE)
cat("BestTune hyperparameters saved as 'Base_Models_Data/Random_Forest/BestHyperparams_RandomForestWithFeatureEngineering.csv'.\n\n")

###########################################
# 3. Make predictions and create confusion matrix CSV with TP, FP, FN, TN
###########################################
# Predict on the test set (for class labels)
predictions <- predict(rf_model, newdata = pca_test)

# Compute the confusion matrix
conf_matrix <- confusionMatrix(predictions, pca_test$label)
print(conf_matrix)

# Extract confusion matrix values; assumes binary classification.
levels_labels <- levels(pca_test$label)
if(length(levels_labels) == 2) {
  positive_class <- levels_labels[1]  # "Drowsy"
  negative_class <- levels_labels[2]  # "Natural"
  
  cm_table <- conf_matrix$table
  TP <- cm_table[positive_class, positive_class]
  FP <- cm_table[positive_class, negative_class]
  FN <- cm_table[negative_class, positive_class]
  TN <- cm_table[negative_class, negative_class]
  
  cm_df <- data.frame(model = "RandomForestWithFeatureEngineering",
                      TP = TP,
                      FP = FP,
                      FN = FN,
                      TN = TN)
  
  write.csv(cm_df, "Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestWithFeatureEngineering.csv", row.names = FALSE)
  cat("Confusion matrix saved as 'Base_Models_Data/Random_Forest/ConfusionMatrix_RandomForestWithFeatureEngineering.csv'.\n\n")
  
  ###########################################
  # 4. Calculate and save F1-score, Precision, Recall, Accuracy, and AUC as CSV
  ###########################################
  accuracy  <- conf_matrix$overall["Accuracy"]
  recall    <- conf_matrix$byClass["Sensitivity"]
  precision <- conf_matrix$byClass["Pos Pred Value"]
  f1_score  <- 2 * (precision * recall) / (precision + recall)
  
  # Obtain probability predictions for ROC curve (type = "prob")
  probs <- predict(rf_model, newdata = pca_test, type = "prob")
  # Use probability for the positive class ("Drowsy")
  roc_obj <- roc(response = pca_test$label, predictor = probs[[positive_class]])
  auc_val <- as.numeric(auc(roc_obj))
  
  metrics_df <- data.frame(model = "RandomForestWithFeatureEngineering",
                           Accuracy = accuracy,
                           Precision = precision,
                           Recall = recall,
                           FOne = f1_score,
                           AUC = auc_val)
  
  write.csv(metrics_df, "Base_Models_Data/Random_Forest/Metrics_RandomForestWithFeatureEngineering.csv", row.names = FALSE)
  cat("Performance metrics saved as 'Base_Models_Data/Random_Forest/Metrics_RandomForestWithFeatureEngineering.csv'.\n")
  
  ###########################################
  # 5. Create and save an AUC ROC plot as a PNG
  ###########################################
  # Build ROC data: fpr (1 - specificity) and tpr (sensitivity)
  roc_data <- data.frame(fpr = 1 - roc_obj$specificities, tpr = roc_obj$sensitivities)
  
  write.csv(roc_data, "Base_Models_Data/Random_Forest/ROCData_RandomForestWithFeatureEngineering.csv", row.names = FALSE)
  cat("ROC data saved as 'Base_Models_Data/Random_Forest/ROCData_RandomForestWithFeatureEngineering.csv'.\n")
  
  # Generate the ROC plot using ggplot2
  roc_plot <- ggplot(roc_data, aes(x = fpr, y = tpr)) +
    geom_line(color = "blue") +  # ROC curve in blue
    geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +  # 45° dotted reference line
    labs(x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         title = "ROC Curve for RandomForestWithFeatureEngineering") +
    theme_minimal() +
    # Annotate AUC in the top right corner (rounded to 3 decimals)
    annotate("text", x = 0.75, y = 0.95, 
             label = paste("AUC =", format(round(auc_val, 3), nsmall = 3)), 
             color = "red", size = 5)
  
  # Save the ROC plot as a PNG file
  ggsave("Base_Models_Data/Random_Forest/ROC_RandomForestWithFeatureEngineering.png", 
         plot = roc_plot, width = 7, height = 7, dpi = 300)
  cat("ROC plot saved as 'Base_Models_Data/Random_Forest/ROC_RandomForestWithFeatureEngineering.png'.\n")
  
} else {
  cat("Skipping confusion matrix extraction and ROC analysis: classification is not binary.\n")
}

```
{{< pagebreak >}}

# References

::: {#refs}
:::

{{< pagebreak >}}

# Code Appendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
