---
title: "Stat 380 Final"
author: "Patrick Erickson, Aiden Shindel, Jordan Brophy"
format: 
  pdf:
    cite-method: biblatex
    pdf-engine: xelatex  
editor: visual
bibliography: references.bib
csl: mla.csl
nocite: |
  @*
appendix: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

# Predicting Drowsy Driving: Can we Machine Learning Model to be able to Predict Drowsiness in Drivers?

## Introduction

### Intuition Behind the Choice of Data Set

-   The data set must be large to be able to get a good testing pool.
-   The data set has meaningful predictors are practically relevant to our test case.
-   The data set contains a drowsy/awake variable
-   The data set follows **F.A.I.R.** and **C.A.R.E.** principles.

### Ethical Considerations in the Choice of our Data Set

In order to ensure the ethical feasibility of the data set, we also scrutinized our selection to follow the **F.A.I.R.** and **C.A.R.E.** principles as discussed in class.

Since we had obtained our data from the open-source data hub Kaggle, we can ensure that anyone can **find** and **access** this data set through Kaggle's use guidelines, stating for a free distribution of data sets that are published for public use to practice ML or data analysis on. The data is also highly versatile, taking a csv format. This is one of the most **interoperable** formats there are for data sets, especially for our use-case in R, where reading in a csv is already built into the base. Lastly, the data set can be **reused** for multiple types of data analysis, and will always remain relevant for its time period.

In terms of **collective benefit**, the data set contributes to an overall understanding of driving without harming any individuals , due to the fact that it is simply raw data. Secondly, as I had stated the author's name, there is clearly ownership over the data set coming from kaggle as mentioned, showing proper **Authority to Control**. Since proper attributions are maintained, it also falls under the **Responsibility and Ethics** guidelines of the C.A.R.E. principles.

The dataset can be found [here](https://www.kaggle.com/datasets/yasharjebraeily/drowsy-detection-dataset?select=Drowsy_datset).

### Cleaning

**Aiden, Jordan, Here are some notes for yall. We gotta make sure we handling pulling PNGs and JPEG (not jpg).**


```{r}
#|Warning: FALSE
library(png)
library(jpeg)
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(knitr)
library(purrr)
library(BiocManager)
if (!requireNamespace("EBImage", quietly = TRUE)) {
  BiocManager::install("EBImage")
}
library(EBImage)
```


Note that the dataset name is Drowsy_datset

```{r}
train_drowsy_dir <- "Drowsy_datset/train/DROWSY"
train_natural_dir <- "Drowsy_datset/train/NATURAL"
test_drowsy_dir <- "Drowsy_datset/test/NATURAL"
test_natural_dir <- "Drowsy_datset/test/NATURAL"

# Check files in each directory
drowsy_files_train <- list.files(train_drowsy_dir, pattern = "\\.png$", full.names = TRUE)
natural_files_train <- list.files(train_natural_dir, pattern = "\\.png$", full.names = TRUE)
drowsy_files_test <- list.files(test_drowsy_dir, pattern = "\\.png$", full.names = TRUE)
natural_files_test <- list.files(test_natural_dir, pattern = "\\.png$", full.names = TRUE)

# Assuming drowsy_files and natural_files are already defined
# For DROWSY files:

# For DROWSY files: Save the first 10 file names in a data frame.
df_drowsy_train <- data.frame(File = head(drowsy_files_train, 5), stringsAsFactors = FALSE)
df_natural_train <- data.frame(File = head(natural_files_train, 5), stringsAsFactors = FALSE)
df_drowsy_test <- data.frame(File = head(drowsy_files_test, 5), stringsAsFactors = FALSE)
df_natural_test <- data.frame(File = head(natural_files_test, 5), stringsAsFactors = FALSE)

# Print the data frames (since they only have 5 rows, printing them prints the head)
kable(df_natural_train, caption = "First 5 NATURAL Train Files:")
kable(df_drowsy_train, caption = "First 5 DROWSY Train Files:")
kable(df_natural_test, caption = "First 5 NATURAL Test Files:")
kable(df_drowsy_test, caption = "First 5 DROWSY Test Files:")

```

This is showing that some files are PNG and some are JPEG.

```{r}
check_png <- function(filepath) {
  tryCatch({
    img <- png::readPNG(filepath)
    return(TRUE)
  }, error = function(e) {
    return(FALSE)
  })
}

# Check a few files with readJPEG:
for (f in drowsy_files_train[1:5]) {
  cat(f, " => PNG check: ", check_png(f), "\n")
}

check_jpeg <- function(filepath) {
  tryCatch({
    jpeg::readJPEG(filepath)
    return(TRUE)
  }, error = function(e) {
    return(FALSE)
  })
}

# Check a few files with readJPEG:
for (f in drowsy_files_train[1:5]) {
  cat(f, " => JPEG check: ", check_jpeg(f), "\n")
}

```

We'll have handle BOTH. The following code chunk loads ALL of the images into a drowsy and natural set, by handling png or jpeg (our data cleaning)

```{r}

# Define a wrapper function to read an image
read_image <- function(filepath) {
  if (check_png(filepath)) {
    img <- png::readPNG(filepath)
  } else if (check_jpeg(filepath)) {
    img <- jpeg::readJPEG(filepath)
  } else {
    stop("Unrecognized image format for file: ", filepath)
  }
  return(img)
}

# We use the function to load all the images as vectors
loaded_images_drowsy_train <- lapply(drowsy_files_train, read_image)
loaded_images_natural_train <- lapply(natural_files_train, read_image)
# We use the function to load all the images as vectors
loaded_images_drowsy_test <- lapply(drowsy_files_test, read_image)
loaded_images_natural_test <- lapply(natural_files_test, read_image)
```


We combine the datasets into one train and test dataset.

```{r}
# Create a base data frame using data.frame()
image_dataset_train <- data.frame(
  image = I(c(loaded_images_drowsy_train, loaded_images_natural_train)),
  label = c(rep("Drowsy", length(loaded_images_drowsy_train)), rep("Natural", length(loaded_images_natural_train))),
  stringsAsFactors = FALSE
)

image_dataset_test <- data.frame(
  image = I(c(loaded_images_drowsy_test, loaded_images_natural_test)),
  label = c(rep("Drowsy", length(loaded_images_drowsy_test)), rep("Natural", length(loaded_images_natural_test))),
  stringsAsFactors = FALSE
)
```


We look at the initial class imbalance. Notice its not that bad. **Aiden or Jordan, we could use a class imbalance histogram here**
```{r}
# For the training dataset:
train_label_counts <- image_dataset_train %>% 
  count(label)
print(train_label_counts)

# For the testing dataset:
test_label_counts <- image_dataset_test %>% 
  count(label)
print(test_label_counts)

# slight imbalance, we will use AUC ROC and f1 scores
```

We shuffle the dataset

```{R}

## Shuffle Datasets
set.seed(100)
image_dataset_train <- image_dataset_train[sample(nrow(image_dataset_train)), ]
image_dataset_test <- image_dataset_test[sample(nrow(image_dataset_test)), ]

```

### Preprocessing
We need to ensure that the datasets are the same going in. This applies CLAHE and HFE to apply pre-processing.

```{R}
applyFeatureEngineering <- function(df) {
  
  # Helper function to apply CLAHE using EBImage.
  apply_CLAHE <- function(img_matrix) {
    # Convert the 48x48 matrix into an EBImage object.
    img <- EBImage::Image(img_matrix)
    
    # Apply CLAHE.
    # Note: Your version of EBImage must have the clahe() function available.
    img_clahe <- EBImage::clahe(img, nx = 8, ny = 8, bins = 256, limit = 2, keep.range = FALSE)

    
    # Convert back to a matrix.
    return(as.matrix(img_clahe))
  }
  
  # Helper function for High Frequency Extraction (HFE).
  high_frequency_extraction <- function(img_matrix) {
    # Convert the 48x48 matrix to an EBImage object.
    img <- EBImage::Image(img_matrix)
    
    # Apply a Gaussian blur to capture low frequency details (sigma can be adjusted).
    img_blurred <- EBImage::gblur(img, sigma = 1)
    
    # Subtract the blurred image from the original image to extract high frequencies.
    img_hfe <- img - img_blurred
    
    # Convert the result back to a matrix.
    return(as.matrix(img_hfe))
  }
  
  # Apply the feature engineering functions to each image in the dataset.
  # This creates two new columns:
  #   - image_clahe: with CLAHE-enhanced images.
  #   - image_hfe: with high frequency extracted images.
  df_new <- df %>%
    mutate(
      image = map(image, ~ apply_CLAHE(.))
    ) %>%
    mutate(
      image = map(image, ~ high_frequency_extraction(.))
    )
  
  return(df_new)
}

```

```{r}
processedtrain <- applyFeatureEngineering(image_dataset_train)
processedTest <- applyFeatureEngineering(image_dataset_test)

```


### Readying for models
We define our new models as one with feature engineering and one without feature engineering. We stretch and flatten our images into one long string of vectors.

```{r}
trainNoFeatureEngineering <- image_dataset_train %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")

testNoFeatureEngineering <- image_dataset_test %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")
```

```{r}
trainWithFeatureEngineering <- processedtrain %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")

testWithFeatureEngineering <- processedTest %>%
  mutate(flat_image = map(image, ~ as.vector(t(.)))) %>%
  select(-image) %>%
  unnest_wider(flat_image, names_sep = "_")
```



These are plotting functions.
```{r}
# We'll assume your expanded data frame is called train_df_expanded.
# Each row has a "label" column plus columns flat_image_1, flat_image_2, ..., flat_image_2304.

reconstruct_image <- function(df_row, label_col = "label") {

  label_value <- df_row[[label_col]]
  pixel_columns <- grep("^flat_image_", names(df_row), value = TRUE)
  pixel_values  <- as.numeric(df_row[pixel_columns])
  mat <- matrix(pixel_values, nrow = 48, ncol = 48, byrow = TRUE)
  
  image(
    x      = 1:48,
    y      = 1:48,
    z      = t(apply(mat, 2, rev)),
    col    = gray(seq(0, 1, length.out = 256)),
    main   = paste("Label:", label_value),
    xlab   = "",
    ylab   = "",
    axes   = FALSE,
    asp    = 1
  )
  
  return(mat)
}

```

```{r}

# Function that plots the feature flattened images for 
plot_image_ggplot <- function(df_row, label_col = "label") {
  label_value <- df_row[[label_col]]
  
  pixel_columns <- grep("^flat_image_", names(df_row), value = TRUE)
  pixel_values  <- as.numeric(df_row[pixel_columns])
  
  image_matrix <- matrix(pixel_values, nrow = 48, ncol = 48, byrow = TRUE)
  
  df_image <- expand.grid(x = 1:48, y = 1:48)
  
  df_image$fill <- as.vector(t(image_matrix))
  
  p <- ggplot(df_image, aes(x = x, y = y, fill = fill)) +
    geom_raster() +
    scale_fill_gradient(low = "black", high = "white") +
    scale_y_reverse() +
    labs(title = paste("Label:", label_value)) +
    theme_minimal() +
    theme(
      axis.title = element_blank(),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      plot.title = element_text(hjust = 0.5)
    )
  
  return(p)
}


```


We print the first 9. Notice the differences.
```{r}

# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(trainNoFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)


```

```{r}

# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(testNoFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)



```

```{r}
# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(trainWithFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)
```

```{r}
# Create list of plots for the first 9 images
plot_list <- lapply(1:9, function(i) {
  plot_image_ggplot(testWithFeatureEngineering[i, ])
})

# Arrange and display the plots in a 3x3 grid
grid_arrange_result <- grid.arrange(grobs = plot_list, ncol = 3, nrow = 3)
```

```{r}
write.csv(trainNoFeatureEngineering, file = "Prepped_Data/trainNoFeatureEngineering.csv", row.names = FALSE)
write.csv(testNoFeatureEngineering, file = "Prepped_Data/testNoFeatureEngineering.csv", row.names = FALSE)
write.csv(trainWithFeatureEngineering, file = "Prepped_Data/trainWithFeatureEngineering.csv", row.names = FALSE)
write.csv(testWithFeatureEngineering, file = "Prepped_Data/testWithFeatureEngineering.csv", row.names = FALSE)
```


### Do PCA
we will do PCA. We will use the PCA done on the train set to transform our test set to then evaluate with our models later.
