---
title: "Stat 380 Final"
author: "Patrick Erickson, Aiden Shindel, Jordan Brophy"
format: 
  pdf:
    cite-method: biblatex
    pdf-engine: xelatex  
editor: visual
bibliography: references.bib
csl: mla.csl
nocite: |
  @*
appendix: true
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

# Predicting Drowsy Driving: Can we Machine Learning Model to be able to Predict Drowsiness in Drivers?

## Introduction

### Intuition Behind the Choice of Data Set

-   The data set must be large to be able to get a good testing pool.
-   The data set has meaningful predictors are practically relevant to our test case.
-   The data set contains a drowsy/awake variable
-   The data set follows **F.A.I.R.** and **C.A.R.E.** principles.

### Ethical Considerations in the Choice of our Data Set

In order to ensure the ethical feasibility of the data set, we also scrutinized our selection to follow the **F.A.I.R.** and **C.A.R.E.** principles as discussed in class.

Since we had obtained our data from the open-source data hub Kaggle, we can ensure that anyone can **find** and **access** this data set through Kaggle's use guidelines, stating for a free distribution of data sets that are published for public use to practice ML or data analysis on. The data is also highly versatile, taking a csv format. This is one of the most **interoperable** formats there are for data sets, especially for our use-case in R, where reading in a csv is already built into the base. Lastly, the data set can be **reused** for multiple types of data analysis, and will always remain relevant for its time period.

In terms of **collective benefit**, the data set contributes to an overall understanding of driving without harming any individuals , due to the fact that it is simply raw data. Secondly, as I had stated the author's name, there is clearly ownership over the data set coming from kaggle as mentioned, showing proper **Authority to Control**. Since proper attributions are maintained, it also falls under the **Responsibility and Ethics** guidelines of the C.A.R.E. principles.

The dataset can be found [here](https://www.kaggle.com/datasets/yasharjebraeily/drowsy-detection-dataset?select=Drowsy_datset).

### Cleaning

**Aiden, Jordan, Here are some notes for yall. We gotta make sure we handling pulling PNGs and JPEG (not jpg).**

I had to change the dataset name from Drowsy_datset to Drowsy_dataset

```{r}
#|Warning: FALSE
library(png)
library(dplyr)
library(tidyr)
library(caret)
library(ggcorrplot)
library(ggplot2)
library(kableExtra)
library(knitr)
```


```{r}
train_drowsy_dir <- "Drowsy_datset/train/DROWSY"
train_natural_dir <- "Drowsy_datset/train/NATURAL"

# Check files in each directory
drowsy_files <- list.files(train_drowsy_dir, pattern = "\\.png$", full.names = TRUE)
natural_files <- list.files(train_natural_dir, pattern = "\\.png$", full.names = TRUE)

# Assuming drowsy_files and natural_files are already defined
# For DROWSY files:

# For DROWSY files: Save the first 10 file names in a data frame.
df_drowsy <- data.frame(File = head(drowsy_files, 5), stringsAsFactors = FALSE)
df_natural <- data.frame(File = head(natural_files, 5), stringsAsFactors = FALSE)

# Print the data frames (since they only have 10 rows, printing them prints the head)
kable(df_natural, caption = "First 5 NATURAL Files:")
kable(df_drowsy, caption = "First 5 DROWSY Files:")

```

This is showing that some files had extra attachments to them. We need to remove the .jpeg so that they are all one format.


```{r}
library(jpeg)
check_png <- function(filepath) {
  tryCatch({
    img <- png::readPNG(filepath)
    return(TRUE)
  }, error = function(e) {
    return(FALSE)
  })
}

# Check a few files with readJPEG:
for (f in drowsy_files[1:5]) {
  cat(f, " => PNG check: ", check_png(f), "\n")
}


library(jpeg)

check_jpeg <- function(filepath) {
  tryCatch({
    jpeg::readJPEG(filepath)
    return(TRUE)
  }, error = function(e) {
    return(FALSE)
  })
}

# Check a few files with readJPEG:
for (f in drowsy_files[1:5]) {
  cat(f, " => JPEG check: ", check_jpeg(f), "\n")
}

```
We'll have handle BOTH.

```{r}


```
